---
title: 'Predicting NFL Wide Receiver Fantasy Performance in their Rookie Season'
author: "Dash Beavers, Shiley Einav, Zach Teplin"
date: "6/7/23"
output:
  html_document:
    df_print: paged
header-includes:
- \usepackage{bbm, amsmath,amsfonts,amsthm,amssymb,mathrsfs,amsxtra,amscd,latexsym,
  xcolor, graphicx, fvextra}
- \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(knitr)) #makes pdfs
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(library(ggplot2)) #makes nice plots
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(glmnet))
suppressPackageStartupMessages(library(predtools))
suppressPackageStartupMessages(library(psych))
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(library(PerformanceAnalytics))
```

## Loading data files

```{r}

draft_order = read.csv("nfl_draft_wr_2000_2023.csv")

ncaa10 = read.csv("all_data/NCAA Data - 2012.csv")
ncaa9 = read.csv("all_data/NCAA Data - 2013.csv")
ncaa8 = read.csv("all_data/NCAA Data - 2014.csv")
ncaa7 = read.csv("all_data/NCAA Data - 2015.csv")
ncaa6 = read.csv("all_data/NCAA Data - 2016.csv")
ncaa5 = read.csv("all_data/NCAA Data - 2017.csv")
ncaa4 = read.csv("all_data/NCAA Data - 2018.csv")
ncaa3 = read.csv("all_data/NCAA Data - 2019.csv")
ncaa2 = read.csv("all_data/NCAA Data - 2020.csv")
ncaa1 = read.csv("all_data/NCAA Data - 2021.csv")
ncaa0 = read.csv("all_data/NCAA Data - 2022.csv")

combine10 = read.csv("all_data/Combine Data - 2013.csv")
combine9 = read.csv("all_data/Combine Data - 2014.csv")
combine8 = read.csv("all_data/Combine Data - 2015.csv")
combine7 = read.csv("all_data/Combine Data - 2016.csv")
combine6 = read.csv("all_data/Combine Data - 2017.csv")
combine5 = read.csv("all_data/Combine Data - 2018.csv")
combine4 = read.csv("all_data/Combine Data - 2019.csv")
combine3 = read.csv("all_data/Combine Data - 2020.csv")
combine2 = read.csv("all_data/Combine Data - 2021.csv")
combine1 = read.csv("all_data/Combine Data - 2022.csv")
combine0 = read.csv("all_data/Combine Data - 2023.csv")

nfl10 = read.csv("all_data/NFL Fantasy Data - 2013.csv")
nfl9 = read.csv("all_data/NFL Fantasy Data - 2014.csv")
nfl8 = read.csv("all_data/NFL Fantasy Data - 2015.csv")
nfl7 = read.csv("all_data/NFL Fantasy Data - 2016.csv")
nfl6 = read.csv("all_data/NFL Fantasy Data - 2017.csv")
nfl5 = read.csv("all_data/NFL Fantasy Data - 2018.csv")
nfl4 = read.csv("all_data/NFL Fantasy Data - 2019.csv")
nfl3 = read.csv("all_data/NFL Fantasy Data - 2020.csv")
nfl2 = read.csv("all_data/NFL Fantasy Data - 2021.csv")
nfl1 = read.csv("all_data/NFL Fantasy Data - 2022.csv")


```

## Manipulating tables

```{r}

draft_order_0 = filter(draft_order, Year == 2023)
draft_order_1 = filter(draft_order, Year == 2022)
draft_order_2 = filter(draft_order, Year == 2021)
draft_order_3 = filter(draft_order, Year == 2020)
draft_order_4 = filter(draft_order, Year == 2019)
draft_order_5 = filter(draft_order, Year == 2018)
draft_order_6 = filter(draft_order, Year == 2017)
draft_order_7 = filter(draft_order, Year == 2016)
draft_order_8 = filter(draft_order, Year == 2015)
draft_order_9 = filter(draft_order, Year == 2014)
draft_order_10 = filter(draft_order, Year == 2013)

combine0 = filter(combine0, Pos == "WR")
combine1 = filter(combine1, Pos == "WR")
combine2 = filter(combine2, Pos == "WR")
combine3 = filter(combine3, Pos == "WR")
combine4 = filter(combine4, Pos == "WR")
combine5 = filter(combine5, Pos == "WR")
combine6 = filter(combine6, Pos == "WR")
combine7 = filter(combine7, Pos == "WR")
combine8 = filter(combine8, Pos == "WR")
combine9 = filter(combine9, Pos == "WR")
combine10 = filter(combine10, Pos == "WR")

nfl1 = filter(nfl1, FantPos == "WR")
nfl2 = filter(nfl2, FantPos == "WR")
nfl3 = filter(nfl3, FantPos == "WR")
nfl4 = filter(nfl4, FantPos == "WR")
nfl5 = filter(nfl5, FantPos == "WR")
nfl6 = filter(nfl6, FantPos == "WR")
nfl7 = filter(nfl7, FantPos == "WR")
nfl8 = filter(nfl8, FantPos == "WR")
nfl9 = filter(nfl9, FantPos == "WR")
nfl10 = filter(nfl10, FantPos == "WR")

```

## Creating singular data frame of rookie players

```{r}

drafted_ncaa1 = merge(draft_order_1, ncaa1, by = "Player")
drafted_ncaa2 = merge(draft_order_2, ncaa2, by = "Player")
drafted_ncaa3 = merge(draft_order_3, ncaa3, by = "Player")
drafted_ncaa4 = merge(draft_order_4, ncaa4, by = "Player")
drafted_ncaa5 = merge(draft_order_5, ncaa5, by = "Player")
drafted_ncaa6 = merge(draft_order_6, ncaa6, by = "Player")
drafted_ncaa7 = merge(draft_order_7, ncaa7, by = "Player")
drafted_ncaa8 = merge(draft_order_8, ncaa8, by = "Player")
drafted_ncaa9 = merge(draft_order_9, ncaa9, by = "Player")
drafted_ncaa10 = merge(draft_order_10, ncaa10, by = "Player")

full_1 = merge(drafted_ncaa1, nfl1, by = "Player")
colnames(full_1)[9] = "Rk.x"
colnames(full_1)[26] = "Rk.y"

full_2 = merge(drafted_ncaa2, nfl2, by = "Player")
full_3 = merge(drafted_ncaa3, nfl3, by = "Player")
full_4 = merge(drafted_ncaa4, nfl4, by = "Player")
full_5 = merge(drafted_ncaa5, nfl5, by = "Player")
full_6 = merge(drafted_ncaa6, nfl6, by = "Player")
full_7 = merge(drafted_ncaa7, nfl7, by = "Player")
full_8 = merge(drafted_ncaa8, nfl8, by = "Player")
full_9 = merge(drafted_ncaa9, nfl9, by = "Player")
full_10 = merge(drafted_ncaa10, nfl10, by = "Player")

full_combine1 = merge(full_1, combine1, by = "Player")
full_combine2 = merge(full_2, combine2, by = "Player")
full_combine3 = merge(full_3, combine3, by = "Player")
full_combine4 = merge(full_4, combine4, by = "Player")
full_combine5 = merge(full_5, combine5, by = "Player")
full_combine6 = merge(full_6, combine6, by = "Player")
full_combine7 = merge(full_7, combine7, by = "Player")
full_combine8 = merge(full_8, combine8, by = "Player")
full_combine9 = merge(full_9, combine9, by = "Player")
full_combine10 = merge(full_10, combine10, by = "Player")

full = rbind(full_1, full_2, full_3, full_4, full_5, full_6, full_7, full_8, full_9, full_10)

full_combine = rbind(full_combine1, full_combine2, full_combine3, full_combine4, full_combine5,
                     full_combine6, full_combine7, full_combine8, full_combine9, full_combine10)

# Extra filtering to remove NaN
full_combine = filter(full_combine, X40yd > 0)
full_combine = filter(full_combine, Vertical > 0)

```

## Correlation matrix no combine

``` {r}

corr_data = data.matrix(full[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x", "PPR", "FantPt")])
chart.Correlation(corr_data, histogram=FALSE, pch=19)

```


## Correlation matrix with combine

``` {r}

corr_data = data.matrix(full_combine[, c("OverallPick","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","X40yd","Vertical", "PPR")])

#trace("chart.Correlation", edit=T)


chart.Correlation(corr_data, histogram=FALSE, pch=19, cex.labels = 20)

```


## Creating data frame of 2023 drafts for testing new new data

```{r}

drafted_ncaa0 = merge(draft_order_0, ncaa0, by = "Player")
test_data = merge(drafted_ncaa0, combine0, by = "Player")

# Filter combine data
test_data = filter(test_data, X40yd > 0)
test_data = filter(test_data, Vertical > 0)

# Updating column names to match training data for prediction
colnames(test_data)[9] = "Rk.x"
colnames(test_data)[12] = "G.x"
colnames(test_data)[13] = "Rec.x"
colnames(test_data)[14] = "Yds.x"
colnames(test_data)[16] = "TD.x"
colnames(test_data)[17] = "Att.x"
colnames(test_data)[18] = "Yds.1.x"
colnames(test_data)[20] = "TD.1.x"
colnames(test_data)[22] = "Yds.2.x"
colnames(test_data)[24] = "TD.2.x"
colnames(test_data)[25] = "X.9999.x"


```


## Preliminary linear regression model no combine

This allowed us to start playing with model creating before having complete data set with combine to start initializing ideas.

```{r}

y = full$PPR
x1 = full$OverallPick
x2 = full$Yds.x
x3 = full$Rec.x

model_mlr = lm(y ~ x1 + x2 + x3, data = full)
summary(model_mlr)

model_draft_pos = lm(y ~ x1, data = full)
summary(model_draft_pos)

model_rec_yds = lm(y ~ x2, data = full)
summary(model_rec_yds)

```


## Lasso regression model no combine

This allowed us to start playing with model creating before having complete data set with combine to start initializing ideas.

```{r}

y = full$PPR
x = data.matrix(full[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x")])

cv_model = cv.glmnet(x, y, alpha = 1)
best_lambda = cv_model$lambda.min

best_model = glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(best_model)

#find R-Squared
y_predicted <- predict(best_model, s = best_lambda, newx = x)
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)
rsq <- 1 - sse/sst
rsq

```


## Preliminary linear regression model with combine; test on 2023 data included

```{r}

# Initialize variables
y = full_combine$PPR
OverallPick = full_combine$OverallPick
X40yd = full_combine$X40yd
x3 = full_combine$Rec.x
Yds.x = full_combine$Yds.x
TD.x = full_combine$TD.x

# Run different model combinations
model_mlr <- lm(y ~ OverallPick + X40yd + Yds.x + TD.x, data = full_combine)
summary(model_mlr)

#model_draft_pos = lm(y ~ x1, data = full)
#summary(model_draft_pos)

#model_rec_yds = lm(y ~ x2, data = full)
#summary(model_rec_yds)

# Run model prediction
test_data$mlr_prediction = predict(model_mlr, newdata=test_data)

```


## Lasso regression model with combine; test on 2023 data included

```{r}
# Organize training data for lasso
y = full_combine$PPR
x = data.matrix(full_combine[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x", "X40yd", "Vertical")])

cv_model = cv.glmnet(x, y, alpha = 1)
best_lambda = cv_model$lambda.min

best_model = glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(best_model)

#find R-Squared
y_predicted <- predict(best_model, s = best_lambda, newx = x)
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)
rsq <- 1 - sse/sst
rsq

# Lasso prediction
# Make test data to match train data
x_test = data.matrix(test_data[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x", "X40yd", "Vertical")])
test_data$lasso_prediction = predict(best_model, s = best_lambda, newx = x_test)

```

## Run models on a year of data we have fantasy points to see how it performs
## Data frame of 2013-2021 and then test of model performance on 2022 data

```{r}
# Training data frame
full_combine_verify = rbind(full_combine2, full_combine3, full_combine4, full_combine5,
                     full_combine6, full_combine7, full_combine8, full_combine9, full_combine10)

full_combine_verify = filter(full_combine_verify, X40yd > 0)
full_combine_verify = filter(full_combine_verify, Vertical > 0)

# Test data frame
test_data_verify = full_combine1
test_data_verify = filter(test_data_verify, X40yd > 0)
test_data_verify = filter(test_data_verify, Vertical > 0)

# MLR model creation
y = full_combine_verify$PPR
OverallPick = full_combine_verify$OverallPick
X40yd = full_combine_verify$X40yd
x3 = full_combine_verify$Rec.x
Yds.x = full_combine_verify$Yds.x
TD.x = full_combine_verify$TD.x

model_mlr_verify <- lm(y ~ OverallPick + X40yd + Yds.x + TD.x, data = full_combine_verify)

# Lasso model
y = full_combine_verify$PPR
x = data.matrix(full_combine_verify[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x", "X40yd", "Vertical")])
cv_model_verify = cv.glmnet(x, y, alpha = 1)
best_lambda_verify = cv_model_verify$lambda.min

best_model_verify = glmnet(x, y, alpha = 1, lambda = best_lambda_verify)

# Run both models
test_data_verify$mlr_pred = predict(model_mlr_verify, newdata=test_data_verify)

lasso_test_verify = data.matrix(test_data_verify[, c("OverallPick","Rk.x","Rec.x","Yds.x","Avg","TD.x",
                         "Plays","Yds.2.x","Avg.2","TD.2.x", "X40yd", "Vertical")])
test_data_verify$lasso_pred = predict(best_model_verify, s = best_lambda_verify, newx = lasso_test_verify)

# RMSE for both models
RMSE_mlr = mean((test_data_verify$PPR - test_data_verify$mlr_pred)^2)
RMSE_lasso = mean((test_data_verify$PPR - test_data_verify$lasso_pred)^2)

RMSE_mlr
RMSE_lasso

```

